A Wait-free Queue with Poly-logarithmic Step Complexity

### 2022-12-15 podc_paper/introduction.tex
working on related work

A great deal of research in the past 25 years has been devoted to the design of linearizable, lock-free queues,
starting with the classic MS-queue of Michael and Scott \cite{DBLP:conf/podc/MichaelS96}.
Their design uses a singly-linked list with pointers to the front and back nodes.
To dequeue or enqueue an element, the front or back pointer is updated by a 
compare-and-swap (CAS) instruction.
If this CAS fails, the operation must retry.
In the worst case, this means that each successful CAS may cause all other processes to
fail and retry, leading to an amortized step complexity of $\Omega(p)$ per operation.
(To measure amortized step complexity of a lock-free implementation, we consider all possible finite executions
and divide the number of steps in the execution by the number of operations initated in the execution.)
Numerous papers have suggested modifications to the MS-queue \cite{list-some}, but 
all still have $\Omega(p)$ amortized step complexity as a result of
contention on the front and back of the queue.
Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} dubbed this the CAS retry problem.
The same problem occurs in array-based implementations of queues \cite{list-some}.

>>>>

A great deal of research in the past 25 years has been devoted to the design of linearizable, lock-free queues.
As well as being a fundamental data structure, queues are needed in
significant concurrent applications, including OS kernels \cite{MP91}, memory reclamation \cite{?},
and sharing resources or tasks.
The lock-free MS-queue of Michael and Scott \cite{DBLP:conf/podc/MichaelS96} is a classic shared queue implementation.
Their design uses a singly-linked list with pointers to the front and back nodes.
To dequeue or enqueue an element, the front or back pointer is updated by a 
compare-and-swap (CAS) instruction.
If this CAS fails, the operation must retry.
In the worst case, this means that each successful CAS may cause all other processes to
fail and retry, leading to an amortized step complexity of $\Omega(p)$ per operation.
(To measure amortized step complexity of a lock-free implementation, we consider all possible finite executions
and divide the number of steps in the execution by the number of operations initated in the execution.)
Numerous papers have suggested modifications to the MS-queue \cite{list-some}, but 
all still have $\Omega(p)$ amortized step complexity as a result of
contention on the front and back of the queue.
Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} dubbed this the CAS retry problem.
The same problem occurs in array-based implementations of queues \cite{list-some}.

### 2022-12-15 podc_paper/related.tex
working on related work

The MS-queue \cite{DBLP:conf/podc/MichaelS96} has stood the test of time.
A version of it is
included in the standard Java Concurrency Package.  %java.util.concurrent.ConcurrentLinkedQueue
However, as mentioned above, it suffers from the CAS retry problem
which means it is lock-free but not wait-free and has an amortized step complexity
of $\Theta(p)$ per operation.

Moir, Nussbaum, and Shalev~\cite{DBLP:conf/spaa/MoirNSS05} presented a
more sophisticated queue by using the \it{elimination} technique. The
elimination mechanism has the dual purpose of allowing operations to
complete in parallel and reducing contention for the queue. An
Elimination Queue consists of an MS-queue augmented with an
elimination array. Elimination works by allowing opposing pairs of
concurrent operations such as an enqueue and a dequeue to exchange
values when the queue is empty or when concurrent operations can be
linearized to empty the queue. Their algorithm makes it possible for
long-running operations to eliminate an opposing operation. The
empirical evaluation showed the throughput of their work is better
than the MS-queue, but the worst case is still the same; in case there
are $p$ concurrent enqueues, their algorithm is not better than
MS-queue.

Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} tried
to make the MS-queue more parallel by introducing the Baskets
Queue. Their idea is to allow more parallelism by treating the
simultaneous enqueue operations as a basket. Each basket has a time
interval in which all its nodes' enqueue operations overlap. Since the
operations in a basket are concurrent, we can order them in any
way. Enqueues in a basket try to find their order in the basket one by
one using \texttt{CAS} operations. However, like the previous
algorithms, if there are $p$ concurrent enqueue operations in a
basket, the amortized step complexity remains $\Omega(p)$ per
operation.

@@

@@

\subsection{Attiya--Fouren Lower Bound}
Because of the \nf{CAS} retry problem in previous list-based queues
one might guess the $\Omega(p)$ term is inherent in the time
complexity of concurrent queues. Attiya and Fouren gave a lower bound
on amortized complexity of lock-free queues with regard to $c$, the
number of concurrent processes. Their result says if $c$ is
$O(\log\log p)$, any implementation of queues using reads, writes and
conditional operations like \nf{CAS} has $\Omega(c)$ amortized
complexity \cite{DBLP:conf/opodis/AttiyaF17}. The surprising point is
that since their result is about when contention is low, their lower
bound does not contradict our algorithm's complexity and we manage to
reach poly-log time complexity.

>>>>

Michael and Scott \cite{DBLP:conf/podc/MichaelS96} describe the early history of concurrent queue algorithms
in the paper that described their MS-queue \cite{DBLP:conf/podc/MichaelS96}, which
is a lock-free queue that has stood the test of time.
A version of it is
included in the standard Java Concurrency Package.  %java.util.concurrent.ConcurrentLinkedQueue
However, as mentioned above, it suffers from the CAS retry problem
which means it is lock-free but not wait-free and has an amortized step complexity
of $\Theta(p)$ per operation.

A number of papers have described ways to cope with the high contention at the front
and back of the MS-queue.
Moir, Nussbaum, and Shalev~\cite{DBLP:conf/spaa/MoirNSS05} 
added an elimination array that allows an enqueue to pass its value directly
to a concurrent dequeue when the queue is empty or when concurrent operations 
can be linearized to empty the queue.
However, when there are $p$ concurrent enqueues (and no dequeues), the CAS retry problem
is still present.
The baskets queue of
Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} 
is another approach to avoid contention:  concurrent enqueues are grouped into baskets
and operations within a basket can be ordered arbitrarily.
However, if there are $p$ concurrent enqueues in the same basket
the CAS retry problem occurs when they must be ordered into a list using CAS instructions.
Both modifications thus still have $\Omega(p)$ amortized step complexity.

-----

@@

\here{Check:
Turek Shasha Prakash locking without blocking (PODS92).
Prakash Lee Johnson non blocking algorithms for concurrent data structures Tech rept 91-002 U of Florida 1991.
}

@@
